Financialisation of Commodity Markets \\\large{Co-movement Behind-the-Scenes}

### 2024-09-27 docs/paper.tex

By grouping together assets by shared attribute dynamics in common
portfolios, asset pricing techniques--and factor models in
particular--are well suited to study co-movement
\citep{fama_common_1993, carhart_persistence_1997, asness_devil_2013, fama_five_factor_2015, hou_digesting_2015, asness_fact_2015, frazzini_betting_2014, asness_quality_2019}.
Careful, theory-based, attribute selection further enables a refined
analysis of market fundamentals
\citep{schwartz_short_2000, miffre_momentum_2007, gorton_fundamentals_2012, cortazar_commodity_2013, yang_investment_2013, daskalaki_factors_2014, szymanowska_anatomy_2014, fernandez_skewness_2018, bakshi_understanding_2019, boons_basis_2019, sakkas_factor_2020}.
We therefore develop a futures-based asset pricing framework that
includes factors constructed using both theoretical and liquidity
considerations and use this framework to carry out our analysis over
four periods of interest. The first, the pre-financialisation period,
starts in 1997 and is naturally followed by financialisation, spanning
the 2004/2008 period, and the financial crisis and its aftermath,
spanning the 2008/2013 period. The last period spans the 2014/2018
period starting with the beginning of the tapering/end of the US
Quantitative Easing program (QE).\\
Using the weekly price, open interest and market position data described
above we construct market portfolios of individual commodity nearby
futures as well as mimicking portfolios for market, commercial hedging
pressure, term structure and liquidity risk factors in returns. While
our market portfolios are long only, equally weighted portfolios of
particular commodity sets, our other mimicking portfolios for risk
factors are constructed as combinations of two sub-portfolios (legs),
one held long, one held short, which respective constituents are pooled
on a weekly basis according to risk factor specific criteria.\\
For each leg, the position weight for a particular asset for a
particular period is contingent on the value of the corresponding
sorting variable and its rank relative to the other assets in the pool.
The sorting variable values for the asset pool are first rescaled so
that the highest (lowest) value becomes 1 (0). Weights are then
determined pairwise, starting from the extremes with the highest and
lowest figures. A coefficient \(\alpha\) is determined by dividing their
difference (1 in this case) by the mean of the rescaled values. The high
figure is attributed a score of \(\frac{1}{n} \times \alpha\) where
\(n\) is the number of assets in the pool while the score for the low
figure is \(\frac{1/n}{\alpha}\). The operation is repeated for the pair
of assets showing the second highest and second lowest values for the
sorting variable and so on until all the assets in the pool are
attributed a score. The individual asset weights are then determined by
dividing each score by the sum of the scores.\\
The period return for a particular mimicking portfolio is calculated as
the difference between the weighted average return for constituents of
the long portfolio and that for the constituents of the short portfolio.
i.e, return on the mimicking portfolio for risk factor \(i\) (\(f_{i}\))
for period \(t\):

\[r_{t}^{f_{i}}=\sum_{j=1}^{x}w^{long}_{j,t} \times r_{j,t}-\sum_{j=n-x}^{n}w^{short}_{j,t} \times r_{j,t}\]
\(n\equiv\) number of commodities in the set considered for mimicking
portfolio construction.\\
\(x = n \cdot s\).\\
\(s\equiv\) selection threshold (\(\frac{1}{3}\) here).\\
\(w^{*}_{j,t}\equiv\) period \(t\) weight for commodity \(j\) on the
long or short leg.\\
\(r_{j,t}\equiv\) period \(t\) front contract return for commodity
\(j\).\\
\(j\equiv\) commodity rank in the ordered set \(Y_{t}^{f_{i}}\).\\
\(Y_{t}^{f_{i}}\equiv\) period \(t\) ordered set of the commodities
considered for mimicking portfolio construction where the ordering rule
is specific to \(f_{i}\).

@@

We use a time series regression approach where we start with regressions
of the US traded individual commodity weekly front contract returns on
returns to the mimicking portfolio for a particular risk factor where
the portfolio is constructed using the whole cross-section of US traded
commodities. The corresponding \(R^{2}\)s are averaged and the process
is repeated for each of our four risk factors (market, hedging pressure,
term structure and liquidity) and periods (1997/2003, 2004/2008,
2008/2013 and 2014/2018). For the long-short factors the results are
reported independently for models where the regressor is the factor
itself or one of its legs (long vs.~short). The same operation is
repeated for the two (\(p=2\)) and three (\(p=3\)) factor model cases,
where regressors include in turn all combinations of any two and three
risk factors respectively, as well as for the four (\(p=4\)) factor
model case where all the risk factor mimicking portfolios are included.
Results are reported for the one-factor model case with the other
results available upon request.

@@

For each period we also implement the analysis described above over
periods of low and high aggregate CHP (\textbf{CHP}), a notion that we
believe is particularly relevant in the context of financialisation
where issues of co-movement and aggregate market behaviour play a
central role. We define \textbf{CHP} as the cross-sectional average of
individual CHPs for a particular pool of assets. i.e., period \(t\)
\textbf{CHP} for asset pool \(x\):

\[\mathbf{CHP}_{t}^{x}=\frac{1}{n}\sum_{j=1}^{n}CHP_{j,t}\] \(n\equiv\)
number of commodities in set \(x\).\\
\(CHP_{i,t}\equiv\) period \(t\) CHP for commodity \(i\) as defined
above.

We further define aggregate backwardation (\textbf{backwardation}) and
contango (\textbf{contango}) \textbf{CHP} regimes as, for a given
period, \textbf{CHP} levels below and above the period's median
respectively and, in contrast with \citep{hong_what_2012} who aggregate
across sub-sectors, the asset pool we consider for \textbf{CHP}
construction include the whole cross-section of US traded commodities.

In the next part of the analysis, our mimicking portfolios for risk
factors, constructed from the set of all US traded commodities are used
as commodity picking devices on the latter commodity set. For each risk
factor and period, individual US traded commodity front returns are
regressed on returns for the corresponding mimicking portfolio.
Individual commodities are sorted by \(R^{2}\) accordingly from highest
to lowest and the top third (highest \(R^{2}\)s) are selected as
commodity picks for the risk factor for the period. i.e, set of picks
\(P_{t}^{f_{i}}\) for risk factor \(i\) (\(f_{i}\)) over period \(t\):

\[P_{t}^{f_{i}}\equiv\left \{ P_{1}, P_{2}, ..., P_{x} \right \}\]
\(x = n \cdot s\).\\
\(n\equiv\) number of commodities in the set considered for the picking
exercise (twenty-four here).\\
\(s\equiv\) selection threshold (\(\frac{1}{3}\) here).\\
\(P_{j}\equiv\) commodity pick \(j\) for risk factor \(f_{i}\) over
period \(t\)\\
\(j\equiv\) rank in the ordered set \(Y_{t}^{f_{i}}\)\\
\[Y_{t}^{f_{i}}\equiv\left \{ \left \{ \tilde{R_{1}^{2}}, \tilde{R_{2}^{2}}, ..., \tilde{R_{n}^{2}} \right \}, \geq \right \}\]
\(n\equiv\) number of commodities in the set considered for picking
(twenty-four here).\\
\(\tilde{R_{j}^{2}}\equiv\) risk factor \(f_{i}\) returns' explanatory
power on commodity \(j\) nearby futures returns over period \(t\) as
defined above.

For each period we report the proportion of time each pick is held in
the long and (/or) short leg of the corresponding risk factor mimicking
portfolio as well as the average pairwise correlation amongst every set
of picks.

In the last part of the analysis, for each period our mimicking
portfolios for risk factors in returns are constructed using each set of
commodity picks independently as the asset pool for portfolio
construction. A similar time series approach to that described above is
then implemented using these newly formed mimicking portfolios. In this
case, a first set of dependent variables includes the sets of commodity
picks themselves, a second set includes the whole cross-section of US
traded commodities while a third set includes the six UK traded metals
considered in the study.

@@

Table~\ref{tbl-stats-no-regimes} displays descriptive statistics for the
commodity series considered in the study; the results show a clear
pattern of increase in average return during financialisation with an
equally weighted portfolio of the US traded assets showing a mean
returns of 15.9\% significant at the 5\% level vs.~6.9\% with a 10\%
significance in the pre-financialisation period. The increase is even
more pronounced for an equally weighted portfolio of GB traded assets
(metals), 2.47\% (not significant) to 21.8\% (significant at the 10\%
level), which sets out the global nature of the phenomenon. Most
commodities show a higher mean return while five--Lumber (XCME), Natural
gas (XNYM), Palladium (XNYM), Platinum (XNYM) and Nickel-primary
(XLME)--show a lower figure with five significant figures--Crude oil-WTI
(XNYM), Heating oil (XNYM), Copper (XCEC), Copper (XLME) and Tin-refined
(XLME)--against none pre-financialisation. Sugar-\#11 wins the race with
a mean return increasing from -4.6\% to 24.9\% followed by copper with
sizeable increases for both markets: 1.5\% to 28.2\% (XCEC) and 0.5\% to
28.6\% (XLME).\\
The pattern of results for volatilities is similar although the increase
is less pronounced. Both equally weighted portfolios shows an increase,
more pronounced for the GB assets, 10\% to 13.6\% (US) vs.~15.7\% to
26.9\% (GB), while out of the thirty assets, eleven show a drop in
volatility: Oats (XCBT), Lean hogs (XCME), Cocoa (IFUS), Coffee-C
(IFUS), Lumber (XCME), Sugar-\#11 (IFUS), Natural gas (XNYM), Crude
oil-WTI (XNYM), Gasoline (XNYM), Heating oil (XNYM), Palladium (XNYM).
Financialisation seems however to have impacted the commodity complex
beyond these mechanical effects of increased long participation ensuing
from the entry on index type investments as our results suggest next.

The cross-sectional nature of financialisation naturally leads us to
extend the individual hedging pressure paradigm by incorporating
aggregate CHP (\textbf{CHP}), a market wide measure of hedging pressure.
The traditional Keynesian hedging pressure paradigm postulates a
negative relationship between the return on an individual commodity
futures and its hedging pressure. The extended Keynesian hedging
pressure paradigm, in the context of financialisation, should also imply
that returns for individual commodities should be high in periods where
\textbf{CHP} is low and vice et versa. In Table~\ref{tbl-stats-regimes}
the mean returns are shown over high (\textbf{contango}) and low
(\textbf{backwardation}) \textbf{CHP} regimes for each time period.\\
Over the first period eighteen of the twenty-four US commodities have
higher mean returns during periods of \textbf{backwardation} as compared
to \textbf{contango}, with the difference being statistically
significant at the 10\% level for two commodities, at the 5\% level for
one and at the 1\% level for another. For the remaining six commodities
the higher mean return during \textbf{contango} is not statistically
significant for any of them. For the six GB commodities, all metals,
five of them had higher mean returns during phases of
\textbf{backwardation} as compared to \textbf{contango} in the first
period, with one of these differences being statistically significant at
the 10\% level while for the remaining asset (Zinc (XLME)) the higher
mean return during \textbf{contango} is not statistically significant.
An equally weighted portfolio of the twenty-four US commodities had a
mean return of 15.4\% during \textbf{backwardation} phases,
statistically significant at the 1\% level, and a mean return of -1.1\%
during \textbf{contango}, both in the first period, with the difference
between them statistically significant at the 5\% level. The
corresponding figures for an equally weighted portfolio of the GB metals
were 4.4\% and 0.3\% respectively. These results provide broad support
for an aggregate version of the Keynesian hedging pressure hypothesis
and suggest that hedgers in the aggregate were engaged in risk transfer
during this period.\\
Over the second period the pattern essentially reverses. For the
twenty-four US commodities over this period, fourteen have higher mean
returns during phases of \textbf{contango} over \textbf{backwardation}.
Four of the six GB metals achieved higher mean returns over
\textbf{contango} relative to \textbf{backwardation} phases with the
difference being statistically significant for one commodity. The
equally weighted US portfolio had a mean return of 15.7\% over
\textbf{backwardation} against 16.2\% during phases of
\textbf{contango}, both statistically significant at the 10\% level. The
difference was considerably more pronounced for the equally weighted GB
metals portfolio with a mean return of 8\% during \textbf{backwardation}
rising to 35.6\% during \textbf{contango}.

@@

Financialisation was a global phenomenon that, through the channel of
index investment, simultaneously influenced the entire spectrum of
liquid commodities. This sweeping impact spurred the transformation of
commodity markets from a collection of idiosyncratic markets into a
unified asset class with interconnectedness emerging as a crucial aspect
that warrants thorough examination.\\
We set out to address this issue by first examining the average of
weekly pairwise correlations between commodities across the entire
cross-section of the commodity assets considered in the study as well as
by country, sector and sub-sector. Table~\ref{tbl-correlations-periods}
shows the results for the four periods of interest with the results
reported for the entire period as well as for periods of
\textbf{backwardation} and \textbf{contango}. For the entire sample the
average pairwise correlation jumps from 0.07 in the first period to 0.16
over financialisation. These findings indicate higher interdependence
within the full cross-section of commodities starting with the
financialisation period, consistent with the empirical findings in
\citet{fryMcKibbin_evolution_2023} and
\citet{mayer_financialization_2017} as well as with the theoretical
model of \citet{basak_model_2016}. The sector analysis sheds light on
the driving force behind these changes. The US metals see the sharpest
increase in pairwise correlations going from 0.22 over the past period
to 0.56 over financialisation while the three other sectors see a much
smaller percentage increase during financialisation relative to the past
period. These findings thus suggest that the US metals sector was the
driving force behind increased average pairwise correlations during
financialisation, a new observation to the best of our knowledge. The
pattern of correlations is very similar across phases of
\textbf{backwardation} and \textbf{contango} for all the sectors and
sub-sectors over the past and financialisation periods which indicates
that increased commodity interdependence over financialisation was not a
phase driven phenomenon in the same way that individual commodity
returns are, as demonstrated in Table~\ref{tbl-stats-regimes}.\\
We further examine these correlations by year in
Table~\ref{tbl-correlations-years} to uncover any distinct breakpoints.
For the full sample of assets, the average pairwise correlations
increase from 0.08 to 0.12 from 2003 to 2004 seemingly driven by US
metals average pairwise correlations, more than doubling from 0.22 to
0.49, suggesting 2004 as the onset of financialization as noted in
several studies \citep{baker_financialization_2021, tang_index_2012} and
our contribution is to observe that the US metals sub-sector seems to
have played a major role in this.

@@

market factor portfolio. We therefore regress the weekly individual

@@

of 0.80. In contrast, the corresponding figures for the energy sector

@@

explanatory power for the (US based) market factor to almost 12\% over

@@

term structure factors respectively. Here again the dramatic increase in
pricing performance for the CHP factor is driven by the metals sector,

@@

Interestingly, although the factor itself only sees a very modest
increase, its short leg (low roll-yield assets: Working's backwardation)
shows a sizeable rise in pricing performance, yet again driven the
metals sector, with the average \(R^{2}\) for the group's US traded
assets jumping to 28.2\% during financialisation from 6.2\% over the
past period. The pattern of results for both factors extends to the GB
trade assets and suggest a convergence of Keneysian and Working's term
structure perspective over the financialisation period.\\

@@

therefore appear to complement the results of Kang et al.~(2020) who

@@

The crisis period (late 2008 to mid-2013) shows a contrasting pattern of
results. Returns fall sharply relative to the financialisation period
(Table~\ref{tbl-stats-no-regimes}) with eighteen US commodities along
with all six GB traded metals showing lower mean returns. Returns for
equally weighted portfolios of both sets of assets fall down to 9.4\%
and 5.3\% from 15.9\% and 21.8\% respectively. Returns are mostly higher
during phases of \textbf{backwardation} relative to \textbf{contango}
(Table~\ref{tbl-stats-regimes}) with fifteen US assets and five of the
GB traded metals showing higher mean returns in \textbf{backwardation}
and the pattern extending to equally weighted portfolios of the two sets
of assets. Volatility is moderately higher for both sets with fifteen US
commodities and three GB metals showing higher figures and interestingly
eminently higher over phases of \textbf{contango}; out of the US complex
only Sugar-\#11 (IFUS) and none of the GB traded metals show higher
volatility figures over \textbf{backwardation} phases. The symmetry
between the dynamics of CHP and \textbf{CHP}'s explanatory power on
individual commodity returns breaks in the crisis period
(Table~\ref{tbl-time-reg-US-CHP}) with both now seemingly converging;
the US cross-sectional average \(R^{2}\) at the individual level
bouncing back up (22.5\% vs 20.8\%) while that at the aggregate level
rises further (14.7\% vs 7.1\%).

// TODO Add correlation results discussion here.

The performance of the market factor dominates the US commodity complex
over the period with a 50\% increase in mean average \(R^{2}\) to over
30\% (Table~\ref{tbl-regressions-factors}) while that of the CHP factor
decreases and, although increasing, remains modest for the open interest
growth and term structure factors. Taken together, these results
indicate the presence of a systematic factor across the entire
cross-section of the US commodities.

@@

The returns pull-back continues into the post-crisis period
(Table~\ref{tbl-stats-no-regimes}) with twenty-two US and four UK traded
assets showing lower mean returns relative to the crisis while the same
number of US commodities and all UK metals show lower figures relative
to the financialisation period; mean returns for equally weighted
portfolios of US and UK assets drop down to -0.2\% and 2.7\%
respectively. Interestingly volatility also recedes substantially with
twenty US assets and all UK metals showing lower volatility figures
relative to both the crisis and financialisation periods suggesting that
investment outflows from the commodity futures complex have had a lower
impact on volatility than the corresponding inflows witnessed over the
financialisation period. The figures for the equally weighted portfolios
of US and UK assets drop to 9.9\% and 16.2\% respectively, very close to
pre-financialisation levels.\\
The pattern of reversion toward the pre-financialisation Keynesian
paradigm initiated during the crisis continues stronger into the
post-crisis period (Table~\ref{tbl-stats-regimes}) with mean returns
higher during phases of \textbf{backwardation} relative to
\textbf{contango}, more so than over the crisis period with seventeen US
commodities showing higher mean returns over \textbf{backwardation}
phases. The difference is significant at the 5\% level for Lumber (XCME)
and at the 10\% level for Palladium (XNYM) while no difference is
significant for the assets showing higher figures over phases of
\textbf{contango}. The pattern is more pronounced for the UK traded
subset with all six assets showing higher mean returns in
\textbf{backwardation} with the difference significant at the 1\% level
for Nickel-primary (XLME), 5\% level for Aluminium (XLME) and 10\% level
for Zinc (XLME). It also extends to equally weighted portfolios of US
and UK traded assets that both show higher mean returns during
\textbf{backwardation} with the difference significant at the 5\% level
for the UK portfolio and the figure close to the 10\% level significance
threshold for the portfolio of US assets. Volatilities are overall
higher in \textbf{contango} although less so than over the previous
period with seventeen US commodities showing higher figures over
\textbf{contango} against twenty-three in the crisis period. The pattern
extends to three of the six UK traded metals with the difference
significant at the 1\% level for Copper (XLME) and Tin (XLME) and at the
5\% level for Zinc (XLME) while differences are not significant for
Aluminium (XLME), Lead (XLME) and Nickel-primary (XLME) that exhibit
higher volatility over \textbf{backwardation} phases. Equally weighted
portfolios of US and UK traded assets on the other hand both show higher
volatility in \textbf{contango} with the difference significant at the
1\% level for both. Over the financialisation period in contrast
volatilities were predominantly higher in \textbf{backwardation} for US
traded assets with fourteen commodities showing higher volatility
figures in these phases; the post-crisis pattern of results for
volatilities thus likewise shows signs of reversion to
pre-financialisation standards. After showing signs of convergence in
the crisis period the dynamics of the explanatory power of CHP and
\textbf{CHP} part ways (Table~\ref{tbl-time-reg-US-CHP}), seemingly
reverting to the Keynesian paradigm observed pre-financialisation,
further suggesting a return to legacy commodity specific dynamics. The
recovery initiated in the crisis period for that of CHP is confirmed
with a stronger increase in the post-crisis period, 27.2\% vs.~22.5\%,
strongest for the metals sector, 37.3\% vs.~23.7\%, while the energy
sector shows the opposite pattern: 11\% vs.~16\%. In contrast, after
increasing stronger in the crisis, that for \textbf{CHP} falls overall
in the post-crisis period, 7.5\% vs 14.7\%, with the decrease strongest
for the energy sector in a relative sense: 2.5\% vs.~14.8\%.

@@

Taken together the patterns of results over the crisis and post-crisis
periods echo those in \citet{basu_definancialisation} in suggesting that
the exogenous shock delivered by the financial crisis combined with the
accommodative monetary policy that ensued seem to have initiated a
reversion toward pre-financialisation fundamentals in the commodity
futures complex.

>>>>

By grouping together assets by shared attribute dynamics in common
portfolios, asset pricing techniques--and factor models in
particular--are well suited to study co-movement
\citep{fama_common_1993, carhart_persistence_1997, asness_devil_2013, fama_five_factor_2015, hou_digesting_2015, asness_fact_2015, frazzini_betting_2014, asness_quality_2019}.
Careful, theory-based, attribute selection further enables a refined
analysis of market fundamentals
\citep{schwartz_short_2000, miffre_momentum_2007, gorton_fundamentals_2012, cortazar_commodity_2013, yang_investment_2013, daskalaki_factors_2014, szymanowska_anatomy_2014, fernandez_skewness_2018, bakshi_understanding_2019, boons_basis_2019, sakkas_factor_2020}.
We therefore develop a futures-based asset pricing framework that
includes factors constructed using both theoretical and liquidity
considerations. Using the weekly returns, open interest and market
position data described above we construct market portfolios of
individual commodity nearby futures as well as mimicking portfolios for
market, commercial hedging pressure, term structure and liquidity risk
factors in returns. While our market portfolios are long only
combinations of particular commodity sets, our other mimicking
portfolios for risk factors are constructed as combinations of two
sub-portfolios (legs), one held long, one held short, which respective
constituents are pooled on a weekly basis according to risk factor
specific criteria.\\
All portfolios are equally weighted and returns for the two-legged
factors are calculated as the difference between the average return for
constituents of the long leg and that for the constituents of the short.
i.e, return on the mimicking portfolio for risk factor \(i\) (\(f_{i}\))
for period \(t\):

\[r_{t}^{f_{i}} = \frac{1}{x} \left ( \sum_{j=1}^{x}r_{j,t} - \sum_{j=n-x}^{n} r_{j,t} \right )\]
\(n\equiv\) number of commodities in the set considered for mimicking
portfolio construction.\\
\(x = n \cdot s\).\\
\(s\equiv\) selection threshold (\(\frac{1}{3}\) here).\\
\(r_{j,t}\equiv\) period \(t\) front contract return for commodity
\(j\).\\
\(j\equiv\) commodity rank in the ordered set \(Y_{t}^{f_{i}}\).\\
\(Y_{t}^{f_{i}}\equiv\) period \(t\) ordered set of the commodities
considered for mimicking portfolio construction where the ordering rule
is specific to \(f_{i}\).

@@

We use a time series regression approach where individual commodity
weekly front futures contract returns are regressed on returns to the
mimicking portfolio for a particular risk factor with the corresponding
\(\beta\)s and \(R^{2}\)s averaged. For the long-short factors the
\(\beta\)s and \(R^{2}\)s are computed for models where the regressor is
the factor itself as well as both its long and short legs independently
(results available upon request).

For each time period considered in the study, being one of the four
multi-year periods described above or any individual year over the
entire sample time period, we carry out our analysis over the entire
period as well as over phases of low and high aggregate CHP
(\textbf{CHP}) over the period; a notion that we believe is particularly
relevant in the context of financialisation where issues of co-movement
and aggregate market behaviour play a central role. We define
\textbf{CHP} as the cross-sectional average of individual CHPs for a
particular pool of assets. i.e., period \(t\) \textbf{CHP} for asset
pool \(x\): \[\mathbf{CHP}_{t}^{x}=\frac{1}{n}\sum_{j=1}^{n}CHP_{j,t}\]
\(n\equiv\) number of commodities in set \(x\).\\
\(CHP_{i,t}\equiv\) period \(t\) CHP for commodity \(i\) as defined
above.

@@

We further define aggregate backwardation (\textbf{backwardation}) and
contango (\textbf{contango}) \textbf{CHP} regimes as, for a given
period, sub-periods with \textbf{CHP} levels below and above the
period's median respectively and, in contrast with
\citep{hong_what_2012} who aggregate across sub-sectors, our
\textbf{CHP} construction draws on the full cross-section of US-traded
assets considered in the study.

@@

Table~\ref{tbl-stats-combined} displays descriptive statistics for the
commodity series considered in the study; the results show a clear
pattern of increase in average return during financialisation with an
equally weighted portfolio of the US traded assets showing a mean
returns of 15.9\% significant at the 5\% level vs.~6.9\% with a 10\%
significance in the pre-financialisation period. The increase is even
more pronounced for an equally weighted portfolio of GB traded assets
(metals), 2.47\% (not significant) to 21.8\% (significant at the 10\%
level), which sets out the global nature of the phenomenon. Most
commodities show a higher mean return while five--Lumber (XCME), Natural
gas (XNYM), Palladium (XNYM), Platinum (XNYM) and Nickel-primary
(XLME)--show a lower figure with five significant figures--Crude oil-WTI
(XNYM), Heating oil (XNYM), Copper (XCEC), Copper (XLME) and Tin-refined
(XLME)--against none pre-financialisation. Sugar-\#11 wins the race with
a mean return increasing from -4.6\% to 24.9\% followed by copper with
sizeable increases for both markets: 1.5\% to 28.2\% (XCEC) and 0.5\% to
28.6\% (XLME).\\
The pattern of results for volatilities is similar although the increase
is less pronounced. Both equally weighted portfolios shows an increase,
more pronounced for the GB assets, 10\% to 13.6\% (US) vs.~15.7\% to
26.9\% (GB), while out of the thirty assets, eleven show a drop in
volatility: Oats (XCBT), Lean hogs (XCME), Cocoa (IFUS), Coffee-C
(IFUS), Lumber (XCME), Sugar-\#11 (IFUS), Natural gas (XNYM), Crude
oil-WTI (XNYM), Gasoline (XNYM), Heating oil (XNYM), Palladium (XNYM).
Financialisation seems however to have impacted the commodity complex
beyond these mechanical effects of increased long participation ensuing
from the entry on index type investments as our results suggest next.\\
The cross-sectional nature of financialisation naturally leads us to
extend the individual hedging pressure paradigm by incorporating
aggregate CHP (\textbf{CHP}), a market wide measure of hedging pressure.
The traditional Keynesian hedging pressure paradigm postulates a
negative relationship between the return on an individual commodity
futures and its hedging pressure. The extended Keynesian hedging
pressure paradigm, in the context of financialisation, should also imply
that returns for individual commodities should be high in periods where
\textbf{CHP} is low and vice et versa. Table~\ref{tbl-stats-combined}
further shows the mean returns and volatilities for high
(\textbf{contango}) and low (\textbf{backwardation}) \textbf{CHP}
regimes for each time period.\\
Over the first period eighteen of the twenty-four US commodities have
higher mean returns during periods of \textbf{backwardation} as compared
to \textbf{contango}, with the difference being statistically
significant at the 10\% level for two commodities, at the 5\% level for
one and at the 1\% level for another. For the remaining six commodities
the higher mean return during \textbf{contango} is not statistically
significant for any of them. For the six GB commodities, all metals,
five of them had higher mean returns during phases of
\textbf{backwardation} as compared to \textbf{contango} in the first
period, with one of these differences being statistically significant at
the 10\% level while for the remaining asset (Zinc (XLME)) the higher
mean return during \textbf{contango} is not statistically significant.
An equally weighted portfolio of the twenty-four US commodities had a
mean return of 15.4\% during \textbf{backwardation} phases,
statistically significant at the 1\% level, and a mean return of -1.1\%
during \textbf{contango}, both in the first period, with the difference
between them statistically significant at the 5\% level. The
corresponding figures for an equally weighted portfolio of the GB metals
were 4.4\% and 0.3\% respectively. These results provide broad support
for an aggregate version of the Keynesian hedging pressure hypothesis
and suggest that hedgers in the aggregate were engaged in risk transfer
during this period.\\
Over the second period the pattern essentially reverses. For the
twenty-four US commodities over this period, fourteen have higher mean
returns during phases of \textbf{contango} over \textbf{backwardation}.
Four of the six GB metals achieved higher mean returns over
\textbf{contango} relative to \textbf{backwardation} phases with the
difference being statistically significant for one commodity. The
equally weighted US portfolio had a mean return of 15.7\% over
\textbf{backwardation} against 16.2\% during phases of
\textbf{contango}, both statistically significant at the 10\% level. The
difference was considerably more pronounced for the equally weighted GB
metals portfolio with a mean return of 8\% during \textbf{backwardation}
rising to 35.6\% during \textbf{contango}.

@@

Financialisation was a global phenomenon, through the channel of index
investment it simultaneously influenced the entire spectrum of liquid
commodities; this sweeping impact spurred the transformation of
commodity markets from a collection of idiosyncratic markets into a
unified asset class with interconnectedness emerging as a crucial aspect
that warrants thorough examination.\\
We strive to address this issue by first examining the average of weekly
returns pairwise correlations between commodities across the entire
cross-section of the commodity assets considered in the study as well as
by country, sector and sub-sector. Table~\ref{tbl-correlations-periods}
shows the results for the four periods of interest with the results
reported for the entire period as well as for periods of
\textbf{backwardation} and \textbf{contango}. For the entire sample the
average pairwise correlation jumps from 0.07 in the first period to 0.16
over financialisation. This finding indicate higher interdependence
within the full cross-section of commodities starting with the
financialisation period, consistent with the empirical findings in
\citet{fryMcKibbin_evolution_2023} and
\citet{mayer_financialization_2017} as well as with the theoretical
model of \citet{basak_model_2016}. The sector analysis sheds light on
the driving force behind these changes. The US metals see the sharpest
increase in pairwise correlations going from 0.22 over the past period
to 0.56 over financialisation while the three other sectors see a much
smaller percentage increase during financialisation relative to the past
period. These results thus suggest that the US metals sector was the
driving force behind increased average pairwise correlations during
financialisation, a new observation to the best of our knowledge. The
pattern of correlations is very similar across phases of
\textbf{backwardation} and \textbf{contango} for all the sectors and
sub-sectors over the past and financialisation periods which indicates
that increased commodity interdependence over financialisation was not a
phase driven phenomenon in the same way that individual commodity
returns are, as demonstrated in Table~\ref{tbl-stats-combined}.\\
We further examine these correlations by year in
Table~\ref{tbl-correlations-years} to uncover any distinct breakpoints.
For the full sample of assets, the average pairwise correlations
increase from 0.08 to 0.12 from 2003 to 2004 seemingly driven by US
metals average pairwise correlations, more than doubling from 0.22 to
0.49, suggesting 2004 as the onset of financialization as noted in
several studies \citep{baker_financialization_2021, tang_index_2012} and
our contribution is to observe that the US metals sub-sector seems to
have played a major role in this.

@@

market factor portfolio. We accordingly regress the weekly individual

@@

of 0.80. In contrast, the corresponding figure for the energy sector

@@

explanatory power for the (US based) market factor, to almost 12\% over

@@

term structure factors respectively. Here again the increase in pricing
performance for the CHP factor is driven by the metals sector,

@@

Interestingly, although the term structure factor itself only sees a
very modest increase, its short leg (low roll-yield assets: Working's
backwardation) shows a sizeable rise in pricing performance, yet again
driven the metals sector, with the average \(R^{2}\) for the group's US
traded assets jumping to 28.2\% during financialisation from 6.2\% over
the past period. The pattern of results for both factors extends to the
GB trade assets and suggest a convergence of Keneysian and Working's
term structure perspective over the financialisation period.\\

@@

therefore appear to complement the results of \citet{kang_tale_2020} who

@@

The crisis period (late 2008 to mid-2013) shows a contrasting pattern of
results. Returns fall sharply relative to the financialisation period
(Table~\ref{tbl-stats-combined}) with eighteen US commodities along with
all six GB traded metals showing lower mean returns. Returns for equally
weighted portfolios of both sets of assets fall down to 9.4\% and 5.3\%
from 15.9\% and 21.8\% respectively. Returns are mostly higher during
phases of \textbf{backwardation} relative to \textbf{contango} with
fifteen US assets and five of the GB traded metals showing higher mean
returns in \textbf{backwardation} and the pattern extending to equally
weighted portfolios of the two sets of assets. Volatility is moderately
higher for both sets with fifteen US commodities and three GB metals
showing higher figures and interestingly eminently higher over phases of
\textbf{contango}; out of the US complex only Sugar-\#11 (IFUS) and none
of the GB traded metals show higher volatility figures over
\textbf{backwardation} phases. The symmetry between the dynamics of CHP
and \textbf{CHP}'s explanatory power on individual commodity returns
breaks in the crisis period (Table~\ref{tbl-time-reg-US-CHP}) with both
now seemingly converging; the US cross-sectional average \(R^{2}\) at
the individual level bouncing back up (22.5\% vs 20.8\%) while that at
the aggregate level rises further (14.7\% vs 7.1\%).

@@

The average pairwise correlation for the entire cross-section of
commodities considered in the study
(Table~\ref{tbl-correlations-periods}) increases further into the crisis
to 0.28 indicating an intensification of the higher interdependence
within the full cross-section of commodities prompted by
financialisation. The result is consistent with the greater
interdependence between equity and commodity markets over this period
documented in \citet{silvennoinen_financialisation_2013}. The figure for
the US traded section of the metals sector only progresses modestly to
0.59 after peaking at 0.63 in 2006 (Table~\ref{tbl-correlations-years}).
In contrast, the GB traded assets see a sharp increase to over 0.65 in
the crisis starting in 2007 up to a peak of 0.69 in 2010. The
agricultural sector on the other hand shows a substantial rise in the
crisis period with the sharpest increase in 2008 when the average
pairwise correlation for the sector jumps to 0.30. Meanwhile, the energy
sector also shows a notable progression starting in 2007 and peaking at
0.63 in 2008 before levelling back down over the rest of the period. The
onset of the financial crisis seems therefore to have had an across the
board impact on the interdependence of commodities in contrast with
financialisation, much more closely related to the metals sector.

@@

The entire cross-section sees an increase in average regression
coefficient relative to the market factor over the crisis period with
the strongest progression for the GB traded metals followed by the
agricultural sector (Table~\ref{tbl-regressions-index}). The pricing
performance of the factor dominates the US commodity complex over the
period (Table~\ref{tbl-regressions-factors}), showing a 50\% increase in
mean average \(R^{2}\) to over 31\%, while that for both the CHP and
open interest factor decreases and, although increasing, remains modest
for the term structure factors; the increase is strongest for the GB
traded metal assets with an average \(R^{2}\) of almost 34\% for the
period. Taken together, these results indicate the presence of a
systematic factor across the entire cross-section of the US commodities.
For both the CHP and term structure factors, both legs show an increase
in average \(R^{2}\) and while substantially diverging over the
preceding financialisation period the pricing performance for both legs
seems to converge back to a pattern closer to that observed
pre-financialisation.
